---
title: "WIHS Causal Inference for Sequential Treatment"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = F, warning = F)
library(tidyverse)
library(knitr)
options(dplyr.summarise.inform = FALSE)

# read data
labs = read.csv("labsum.csv") # CD4, etc.
f22r = read.csv('f22r.csv') # f22r.csv: new recruit antiviral medication history
drug1 = read.csv('drug1.csv') # drug1.csv: antiviral medications
drugmap = read.csv('drugmap.csv', stringsAsFactors = FALSE) # drug code to classification
el = read.csv("el.csv") # baseline covariates
```


```{r create.case.vis}
# create a dataset called 'case.vis' of CASEID, VISIT, (drug) class, (recoded) id

# subset to HIV + patients who have not taken an antiviral (EMTXDR=2)
# filter out HIV negative patients (maindr != 1)

f22r.f = f22r %>% filter(EMTXDR == 2, MAINDR != 1)
drug1.f = drug1 %>% filter(CASEID %in% f22r.f$CASEID)
# length(unique(drug1.f$CASEID)) # 104 (out of 132) end up being treated

# let's look at these treatment sequences/ how much data do we have here 
case.vis = drug1.f %>% group_by(CASEID, VISIT) %>% 
  summarise(n=n()) %>% mutate(start = min(VISIT)) %>% arrange(CASEID, VISIT)

# now, let's look at the kinds of drugs we observe in these sequences


which.drugs = inner_join(case.vis, drug1.f, by = c("CASEID", "VISIT")) %>%
  left_join(drugmap, by = c("DRUGD1"="Num")) %>% 
  dplyr::select(CASEID, VISIT, DRUGD1, Class)


l = which.drugs %>% group_by(CASEID, VISIT) %>% group_map(~ unique(.x$Class))

# lapply a function 

# l is a character vector with at least two elements
dosage = function(l) {
  
  if(length(l) == 1) {return(l)} else {
  
  if ( "NNRTI" %in% l & "NRTI" %in% l ) {
    return("NRTI + NNRTI")
  } else if ( "NNRTI" %in% l & "PI" %in% l ) {
    return("NNRTI + PI")
  } else if ( "NRTI" %in% l & "PI" %in% l ) {
    return("NRTI + PI")
  } else if ( "NRTI" %in% l & "II" %in% l ) {
    return("NRTI + II")
  } else {return("All")} }
}

case.vis$class = unlist( lapply(l, dosage) )
case.vis$id = sapply(case.vis$CASEID, function(x) which(unique(case.vis$CASEID)==x))

```

### Revised Causal Question

- Time 0: time of first antiretroviral treatment
- Further coarsening of treatment (ART vs. HAART)

**Data** Semiannual visits from a cohort of HIV+ women. Baseline data collected upon study entry and lab data collected at each visit.

**Outcome $Y$**: Time $K+1$ viral load.

**General Question**: What is the distribution of $Y_{K+1}$ under two interventions of interest, $Y^{g^\prime}$ and $Y^{g}$. A causal effect is defined as a function of these two distributions (most commonly, difference in expectations).

**Specific Question**: Let $g^{\prime}$ denote $K$ treatments with HAART and $g$ denote $K$ treatments with ART. When $K=3$, the outcome is the viral load at the end of year 2.

**Problem 1**: Unknown causal structure among time-dependent lab values (viral load, CD4, CD8, HSRAT). 

- *Proposed Solution*: Model each using spike/ slab variable selection to infer causal connections (posterior inclusion probabilities). Average causal effect estimates over posterior, $P(\mathcal{G} \vert \mathbf{D})$.

**Problem 2**: Standard issues - significant amount of missing and censored data. 

- *Proposed Solution*: Sample missing/ censored values 



```{r}
#### Revised question ####
# make time 0 the time of first treatment
case.vis2 = case.vis %>% group_by(CASEID) %>% mutate(t = VISIT - min(VISIT))

# coarsen treatment
case.vis2$class2 = ifelse(case.vis2$class == "NRTI", "ART", "HAART")


ggplot(case.vis2, aes(t, id)) + geom_raster(aes(fill=class2)) + 
  theme_minimal() + ggtitle("Treatment Sequences") + ylab("Patient ID") + xlab("Visit Number")

plot2 = case.vis2 %>% group_by(t, class2) %>% summarise(n = n()) %>% 
  mutate(prop = n/length(unique(case.vis$CASEID)))

ggplot(plot2, aes(x = t, y = prop, fill = class2)) + geom_area() + theme_minimal() + 
  ggtitle("Antiretroviral Therapy Use Among WIHS HIV+ Patients") + 
  ylab("% Receiving Therapy") + xlab("Visit")
```

### Baseline Covariatets

```{r warning=FALSE}

base = left_join(case.vis2, el, by = "CASEID") %>% select(CASEID, VISIT, class2, DOB_EL, RACEEL) %>% 
  group_by(CASEID) %>% slice(1) %>% mutate(age = 2003 - DOB_EL)

base$race = ifelse( base$RACEEL == 2, "White, Hispanic", ifelse(base$RACEEL == 3, "Black, Non-Hispanic", "Other") )

ggplot(base, aes(x = age, fill = class2)) + geom_histogram(bins = 40) + theme_minimal() + 
  ggtitle("Age by Treatment Type")

kable(base %>% group_by(race, class2) %>% summarise(n=n()) %>% pivot_wider(names_from = class2, values_from = n))

# how long have they had HIV? insurance type?
hivstat = read.csv("hivstat.csv")

hivstat.cohort = filter(hivstat, CASEID %in% unique(case.vis2$CASEID))


f = read.csv('f00.csv')

```

30 total ART patients versus 74 total HAART (for first treatment).

### Time Varying Covariates (Possible Mediators)

```{r}
# make dataframe that has the sequence of visits we are interested in for each patient
minv = case.vis2 %>% group_by(CASEID) %>% summarise(`1` = min(VISIT))
minv$`0` = minv$`1` - 1
minv$`2` = minv$`1` + 1
minv$`3` = minv$`1` + 2
minv$`4` = minv$`1` + 3
minv.l = minv %>% pivot_longer(!CASEID, values_to = "VISIT")

labsdat = left_join(minv.l, labs, by = c("CASEID", "VISIT")) %>% group_by(CASEID) %>% arrange(CASEID, VISIT)  %>%
  mutate(t = 1:n()) %>% left_join(case.vis2, by = c("CASEID", "VISIT")) %>%
  select(CASEID, VISIT, VLOAD, CD4N, CD8N, CD3N, HSRAT, t.x, class2)


# let's look at time 0 viral load versus time 1; same with CD4
labsdat.plot1 = filter(labsdat, t.x < 3) %>% group_by(CASEID) %>% 
  mutate(VLOAD.d = c(0,diff(VLOAD)), CD4N.d = c(0, diff(CD4N)), CD8N.d = c(0, diff(CD8N)), 
         CD3N.d = c(0, diff(CD3N)), HSRAT.d = c(0, diff(HSRAT))) %>%
  filter(t.x == 2)

ggplot(labsdat.plot1, aes(x = VLOAD.d, fill = class2)) + 
  geom_histogram(position = "identity", alpha = 0.5, bins = 50) + theme_minimal() + 
  ggtitle("Difference in Viral Load After Initiating Treatment") + xlab("Time 1 - Time 0 Viral Load")

ggplot(labsdat.plot1, aes(x = CD4N.d, fill = class2)) + 
  geom_histogram(position = "identity", alpha = 0.5, bins = 50) + theme_minimal() + 
  ggtitle("Difference in CD4N After Initiating Treatment") + xlab("Time 1 - Time 0 Viral Load")

ggplot(labsdat.plot1, aes(x = CD8N.d, fill = class2)) + 
  geom_histogram(position = "identity", alpha = 0.5, bins = 50) + theme_minimal() + 
  ggtitle("Difference in CD8N After Initiating Treatment") + xlab("Time 1 - Time 0 Viral Load")

ggplot(labsdat.plot1, aes(x = CD3N.d, fill = class2)) + 
  geom_histogram(position = "identity", alpha = 0.5, bins = 50) + theme_minimal() + 
  ggtitle("Difference in CD3N After Initiating Treatment") + xlab("Time 1 - Time 0 Viral Load")

ggplot(labsdat.plot1, aes(x = HSRAT.d, fill = class2)) + 
  geom_histogram(position = "identity", alpha = 0.5, bins = 5) + theme_minimal() + 
  ggtitle("Difference in HSRAT After Initiating Treatment") + xlab("Time 1 - Time 0 Viral Load")


```

Clear decline in viral load after 6 months of treatment (for both ART and HAART). Effects on the other variables are less obvious.

### Missingness


```{r}

np = length(unique(labsdat$CASEID))

tab = labsdat %>% group_by(t.x) %>% summarise(n.cd4 = sum(is.na(CD4N))/np, n.cd8 = sum(is.na(VLOAD))/np)
names(tab) = c("Visit Num", "Prop. Missing CD4", "Prop Missing VLOAD")
kable(tab)

nobs = labsdat %>% group_by(CASEID) %>% summarise(nc = sum(is.na(CD8N))) %>% group_by(nc) %>% summarise(n=n())
# proportion missing for each t = 0,...,3

ggplot(nobs, aes(x=nc, y = n)) + geom_bar(stat = "identity") + theme_minimal() + 
  ggtitle("Number of Patients with x missing Lab values") + xlab("x = Number Missing")

```

### Parametric Assumptions for TVC?

```{r}
labsdat2 = labsdat %>% group_by(CASEID) %>% slice(1)

ggplot(labsdat2, aes(x=VLOAD)) + 
    geom_histogram(aes(y=..density..),      # Histogram with density instead of count on y-axis
                   colour="black", fill="white", bins = 30) +
    geom_density(alpha=.2, fill="#FF6666") + theme_minimal() +  # Overlay with transparent density plot
    ggtitle("Raw Viral Load")


ggplot(labsdat2, aes(x=log(VLOAD))) + 
    geom_histogram(aes(y=..density..),      # Histogram with density instead of count on y-axis
                   colour="black", fill="white", bins = 30) +
    geom_density(alpha=.2, fill="#FF6666") + theme_minimal() +  # Overlay with transparent density plot
    ggtitle("Log Viral Load")

```

```{r}
ggplot(labsdat2, aes(x=CD4N)) + 
    geom_histogram(aes(y=..density..),      # Histogram with density instead of count on y-axis
                   colour="black", fill="white", bins = 30) +
    geom_density(alpha=.2, fill="#FF6666") + theme_minimal() +  # Overlay with transparent density plot
    ggtitle("CD4 Cell Count")

```

A CD4 count is a blood test to check the amount of CD4 cells in the body. CD4 cells are a type of white blood cell. They alert other immune cells to the presence of infections such as bacteria and other viruses in the body. CD4 cells are also a subset of immune cells called T cells. When a person is living with HIV, the virus attacks the CD4 cells in their blood. This process damages CD4 cells and causes the number of them in the body to drop, making it difficult to fight infections. A healthy immune system normally has a CD4 count ranging from 500 to 1,600 cells per cubic millimeter of blood (cells/mm3),

```{r}
ggplot(labsdat2, aes(x=CD8N)) + 
    geom_histogram(aes(y=..density..),      # Histogram with density instead of count on y-axis
                   colour="black", fill="white", bins = 30) +
    geom_density(alpha=.2, fill="#FF6666") + theme_minimal() +  # Overlay with transparent density plot
    ggtitle("CD8 Cell Count")
```

The absolute number of all CD8 cells, which include both killer and suppressor T cells. The normal range for an HIV-negative person is 150 to 1,000. It is usually higher in a person with HIV. 

```{r}
ggplot(labsdat2, aes(x=HSRAT)) + 
    geom_histogram(aes(y=..density..),      # Histogram with density instead of count on y-axis
                   colour="black", fill="white", bins = 30) +
    geom_density(alpha=.2, fill="#FF6666") + theme_minimal() +  # Overlay with transparent density plot
    ggtitle("HSRAT: CD4/ CD8")
```


The CD4/CD8 ratio is often used as an indicator of immune system health. The normal range for HIV-negative people is around 1.0 to 4.0, or one to four CD4 cells for every CD8 cell. People with uncontrolled HIV may have fewer CD4 cells than CD8 cells, indicated by a ratio less than 1.0.

```{r}
# labsdat: caseid, visit, vload, cd4n, cd8n, cd3n, hsrat

# let's filter to people who have an observed outcome (time 4)
outcome.cohort = labsdat %>% filter(t.x==4) %>% filter(!is.na(VLOAD))
labsdat.f = labsdat %>% filter(t.x < 5, CASEID %in% outcome.cohort$CASEID)




labsdat.kplus1 = labsdat %>% filter(t.x==4)
nrow(labsdat.kplus1) - sum(is.na(labsdat.kplus1$VLOAD))
```



